---
title: "DATA 624 Project data test"
author: "Samuel I Kigamba"
date: "June 23, 2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


```{r, message=FALSE, warning=FALSE}
library(tidyverse)
library(fpp2)
library(readxl)
library(tseries)
```

# This analysis covers "S02Var03" "S03Var05" & "S03Var07"

```{r}
df <- read.csv("https://raw.githubusercontent.com/dmoste/DATA624/main/Project1/raw_data.csv", header = TRUE)
head(df)

```

```{r}
raw = df[c("SeriesInd", "S02Var03", "S03Var05", "S03Var07")]
summary(raw)

```

S03VAr05 and S03VAr07 looks almost identical from the summary above.

Visualize the data and look for seasonality or trend within the different data sets of data.

```{r}

# Data under S02var03 has an outlier which we eliminate before creating visualizations.
raw %>%
  filter(S02Var03 < 20) %>%
  ggplot(aes(x = SeriesInd, y = S02Var03)) +
    geom_line() +
  ggtitle("S02Var03")

raw %>%
  ggplot(aes(x = SeriesInd, y = S03Var05)) +
    geom_line() +
  ggtitle("S03Var05")

raw %>%
  ggplot(aes(x = SeriesInd, y = S03Var07)) +
    geom_line() +
  ggtitle("S03Var07")

```



```{r}
# Convert to time series

SeriesInd.ts <- ts(raw$SeriesInd)
S02Var03.ts <- na.remove(ts(raw$S02Var03, frequency = 253))
S03Var05.ts <- na.remove(ts(raw$S03Var05, frequency = 253))
S03Var07.ts <- na.remove(ts(raw$S03Var07, frequency = 253))

```

Decompose data and check for trend and seasonality

```{r}
S02Var03.ts %>%
  decompose(type="multiplicative")%>%
  autoplot() + xlab("Time") +
  ggtitle("Decomposition of S02Var03")

S03Var05.ts %>%
  decompose(type="multiplicative") %>%
  autoplot() + xlab("Time") +
  ggtitle("Decomposition of S03Var05")


S03Var07.ts %>%
  decompose(type="multiplicative") %>%
  autoplot() + xlab("Time") +
  ggtitle("Decomposition of S03Var07")

```

None of the three variables exhibits obvious seasonality or cyclicity. S02Var03 has an otlier that seems to compress the graph downwards.


Evaluating performance of simple models using cross-validation and RMSE:

```{r}
S02Var03_rmse_rwf_nodrift <- tsCV(S02Var03.ts, rwf, drift = FALSE, h = 1)
S02Var03_rmse_rwf_nodrift <- sqrt(mean(S02Var03_rmse_rwf_nodrift^2, na.rm = TRUE))
S02Var03_rmse_rwf_drift <- tsCV(S02Var03.ts, rwf, drift = TRUE, h = 1)
S02Var03_rmse_rwf_drift <- sqrt(mean(S02Var03_rmse_rwf_drift^2, na.rm = TRUE))
S02Var03_rmse_meanf <- tsCV(S02Var03.ts, meanf, h = 1)
S02Var03_rmse_meanf <- sqrt(mean(S02Var03_rmse_meanf^2, na.rm = TRUE))

S03Var05_rmse_rwf_nodrift <- tsCV(S03Var05.ts, rwf, drift = FALSE, h = 1)
S03Var05_rmse_rwf_nodrift <- sqrt(mean(S03Var05_rmse_rwf_nodrift^2, na.rm = TRUE))
S03Var05_rmse_rwf_drift <- tsCV(S03Var05.ts, rwf, drift = TRUE, h = 1)
S03Var05_rmse_rwf_drift <- sqrt(mean(S03Var05_rmse_rwf_drift^2, na.rm = TRUE))
S03Var05_rmse_meanf <- tsCV(S03Var05.ts, meanf, h = 1)
S03Var05_rmse_meanf <- sqrt(mean(S03Var05_rmse_meanf^2, na.rm = TRUE))

S03Var07_rmse_rwf_nodrift <- tsCV(S03Var07.ts, rwf, drift = FALSE, h = 1)
S03Var07_rmse_rwf_nodrift <- sqrt(mean(S03Var07_rmse_rwf_nodrift^2, na.rm = TRUE))
S03Var07_rmse_rwf_drift <- tsCV(S03Var07.ts, rwf, drift = TRUE, h = 1)
S03Var07_rmse_rwf_drift <- sqrt(mean(S03Var07_rmse_rwf_drift^2, na.rm = TRUE))
S03Var07_rmse_meanf <- tsCV(S03Var07.ts, meanf, h = 1)
S03Var07_rmse_meanf <- sqrt(mean(S03Var07_rmse_meanf^2, na.rm = TRUE))

```

```{r}

S02Var03_rmse_rwf_nodrift
S02Var03_rmse_rwf_drift
S02Var03_rmse_meanf

S03Var05_rmse_rwf_nodrift
S03Var05_rmse_rwf_drift
S03Var05_rmse_meanf

S03Var07_rmse_rwf_nodrift
S03Var07_rmse_rwf_drift
S03Var07_rmse_meanf

```

From the RMSE analysis of three models, namely: rwf with no drift, rmf with draft and mean, the rfw is the slightly better performing model.

We will use time series cross-validation to compare the one-step forecast accuracy using Simple Exponential Smoothing, Holts Linear trend method and Holds dumped trend methods.

# S02Var03

```{r}
S02Var03_ses <- tsCV(S02Var03.ts, ses, h=140)
S02Var03_holt <- tsCV(S02Var03.ts, holt, h=140)
S02Var03_holtdmpd <- tsCV(S02Var03.ts, holt, damped=TRUE, h=140)

print("S02Var03")
print("MSE")

# Compare MSE:

print(c("ses:", mean(S02Var03_ses^2, na.rm=TRUE)))

print(c("Holt:", mean(S02Var03_holt^2, na.rm=TRUE)))

print(c("Holt damped:", mean(S02Var03_holtdmpd^2, na.rm=TRUE)))


print("MAE")
# Compare MAE:

print(c("ses:", mean(abs(S02Var03_ses), na.rm=TRUE)))

print(c("Holt:", mean(abs(S02Var03_holt), na.rm=TRUE)))

print(c("Holt damped:", mean(abs(S02Var03_holtdmpd), na.rm=TRUE)))

```


Simple Exponential Smoothing method is best whether you compare MAE or MSE values. 
So we will proceed with using the SES method and apply it to the whole data set to get forecasts for future periods.
We will also include the three


```{r}

S02Var03.ts <- raw %>%
  filter(S02Var03 < 20) %>%
  select(S02Var03) %>%
  ts()

# Random walk forecasts without drift
S02Var03_rwf <- rwf(S02Var03.ts, h = 140)
#S02Var03_drwf <- rwf(S02Var03.ts, h = 140, drift = TRUE)
#S02Var03_mean <- meanf(S02Var03.ts, h = 140)

# Exponential Smoothing forecast
S02Var03_ses <- ses(S02Var03.ts, h = 140)

autoplot(S02Var03.ts) +
  autolayer(S02Var03_rwf, series = "Naive", PI = FALSE) +
  #autolayer(S02Var03_drwf, series = "Drift", PI = FALSE) +
  #autolayer(S02Var03_mean, series = "Mean", PI = FALSE) +
  autolayer(S02Var03_ses, series = "Ses", PI = FALSE)

```

Both simple exponential smoothing and naive random walk methods provide identical forcasts.

# Exponential Smoothing


```{r}

S02Var03_ses <- ses(S02Var03.ts, h = 140)
S02Var03_ses[["model"]]

```

The value of alpha of $\alpha = 0.9999$ is very close to one, showing that the level reacts strongly to each new observation. This also makes the ses method almost indistinguishable fom random walk forecast.


```{r}

# Estimate parameters

summary(S02Var03_ses)

```



# S03Var05

```{r}

S03Var05_ses <- tsCV(S03Var05.ts, ses, h=140)
S03Var05_holt <- tsCV(S03Var05.ts, holt, h=140)
S03Var05_holtdmpd <- tsCV(S03Var05.ts, holt, damped=TRUE, h=140)


print("S03Var05")
print("MSE")

# Compare MSE:

print(c("ses:", mean(S03Var05_ses^2, na.rm=TRUE)))

print(c("Holt:", mean(S03Var05_holt^2, na.rm=TRUE)))

print(c("Holt damped:", mean(S03Var05_holtdmpd^2, na.rm=TRUE)))


print("MAE")

# Compare MAE:

print(c("ses:", mean(abs(S03Var05_ses), na.rm=TRUE)))

print(c("Holt:", mean(abs(S03Var05_holt), na.rm=TRUE)))

print(c("Holt damped:", mean(abs(S03Var05_holtdmpd), na.rm=TRUE)))

```


Simple Exponential Smoothing method is best whether you compare MSE values.
Holt damped is slightly better than ses for MAE value but the improvement is not high enough to justify the added complexity. 
So we will proceed with using the SES method and apply it to the whole data set to get forecasts for future periods.


```{r}

# Random walk forecasts without drift
S03Var05_rwf <- rwf(S03Var05.ts, h = 140)
#S03Var05_drwf <- rwf(S03Var05.ts, h = 140, drift = TRUE)
#S03Var05_mean <- meanf(S03Var05.ts, h = 140)

# Exponential Smoothing Forecast
S03Var05_ses <- ses(S03Var05.ts, h = 140)

autoplot(S03Var05.ts) +
  autolayer(S03Var05_rwf, series = "Naive", PI = FALSE) +
  #autolayer(S03Var05_drwf, series = "Drift", PI = FALSE) +
  #autolayer(S03Var05_mean, series = "Mean", PI = FALSE) +
  autolayer(S03Var05_ses, series = "Ses", PI = FALSE)

```

Both simple exponential smoothing and naive random walk methods provide identical forcasts.


# Exponential Smoothing: Check for alpha

```{r}

S03Var05_ses <- ses(S03Var05.ts, h = 140)
S03Var05_ses[["model"]]

```

The value of alpha of $\alpha = 0.8471$ is very close to one, showing that the level reacts strongly to each new observation. This also makes the ses method almost indistinguishable fom random walk forecast.


```{r}

# Estimate parameters

summary(S03Var05_ses)

```




# S03Var07

```{r}

S03Var07_ses <- tsCV(S03Var07.ts, ses, h=140)
S03Var07_holt <- tsCV(S03Var07.ts, holt, h=140)
S03Var07_holtdmpd <- tsCV(S03Var07.ts, holt, damped=TRUE, h=140)


print("S03Var07")
print("MSE")

# Compare MSE:

print(c("ses:", mean(S03Var07_ses^2, na.rm=TRUE)))

print(c("Holt:", mean(S03Var07_holt^2, na.rm=TRUE)))

print(c("Holt damped:", mean(S03Var07_holtdmpd^2, na.rm=TRUE)))


print("MAE")
# Compare MAE:

print(c("ses:", mean(abs(S03Var07_ses), na.rm=TRUE)))

print(c("Holt:", mean(abs(S03Var07_holt), na.rm=TRUE)))

print(c("Holt damped:", mean(abs(S03Var07_holtdmpd), na.rm=TRUE)))


```


Simple Exponential Smoothing method is best whether you compare MSE values.
Holt damped is slightly better than ses for MAE value but the improvement is not high enough to justify the added complexity. 
So we will proceed with using the SES method and apply it to the whole data set to get forecasts for future periods.


```{r}

# Random walk forecasts without drift
S03Var07_rwf <- rwf(S03Var07.ts, h = 140)
#S03Var07_drwf <- rwf(S03Var07.ts, h = 140, drift = TRUE)
#S03Var07_mean <- meanf(S03Var07.ts, h = 140)

# Exponential Smoothing Forecast
S03Var07_ses <- ses(S03Var07.ts, h = 140)

autoplot(S03Var07.ts) +
  autolayer(S03Var07_rwf, series = "Naive", PI = FALSE) +
  #autolayer(S03Var07_drwf, series = "Drift", PI = FALSE) +
  #autolayer(S03Var07_mean, series = "Mean", PI = FALSE) +
  autolayer(S03Var07_ses, series = "Ses", PI = FALSE)

```

Both simple exponential smoothing and naive random walk methods provide identical forcasts with ses having a slight edge.


# Exponential Smoothing: Check for alpha

```{r}

S03Var07_ses <- ses(S03Var07.ts, h = 140)
S03Var07_ses[["model"]]

```

The value of alpha of $\alpha = 0.9999$ is very close to one, showing that the level reacts strongly to each new observation. This also makes the ses method almost indistinguishable fom random walk forecast.


```{r}

# Estimate parameters


summary(S03Var07_ses)

```