---
title: "Data 624 HW1"
author: "Samuel I Kigamba"
date: "June 23, 2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Required Packages

```{r, message=FALSE, warning=FALSE}
library(forecast)
library(ggplot2)
library(seasonal)
library(fma)
library(mlbench)
library(corrplot)
library(caret)
library(e1071)
library(mice)

```
# HW 6.2 Questions

The plastics data set consists of the monthly sales (in thousands) of product A for a plastics manufacturer for five years.

a) Plot the time series of sales of product A. Can you identify seasonal fluctuations and/or a trend-cycle?
b) Use a classical multiplicative decomposition to calculate the trend-cycle and seasonal indices.
c) Do the results support the graphical interpretation from part a?
d) Compute and plot the seasonally adjusted data.
e) Change one observation to be an outlier (e.g., add 500 to one observation), and recompute the seasonally adjusted data. What is the effect of the outlier?
f) Does it make any difference if the outlier is near the end rather than in the middle of the time series?

```{r}
p = plastics
print(p)

```

# Solutions:

# Part A: Plot the time series of sales of product A. Can you identify seasonal fluctuations and/or a trend-cycle?

```{r}
p <- plastics
autoplot(p, series="Data") + xlab("Year") + ylab("Sales") + ggtitle("Product A: Monthly Sales '000'")

```

Conclusion: The plot above shows that there is a seasonal nature to the plastics data set. It appears that the seasonality is annual will sales peaking at mid year. 
There trend cycle appears to be positive with strong sales growth every year.


# Part B: Use a classical multiplicative decomposition to calculate the trend-cycle and seasonal indices.


```{r}

# Decompose plastics data
m = decompose(p, type="multiplicative")

# Calculating trend-cycle
print(m$trend)

# Calculating seasonal indices
print(m$seasonal)

# Calculating remainder component
print(m$random)

```

Conclusion: See output above.
Note that the estimate of the trend-cycle is unavailable for the first 6 and last 6 observations and as a result there is also no estimate of the remainder component for the same period.
This is a common problem with the classical decomposition.

# Part C: Do the results support the graphical interpretation from part a?


```{r}

# Plot seasonal indices
autoplot(m$seasonal)

# plot trend-cycle
autoplot(m$trend)

# plot trend-cycle
autoplot(m$random)

```

Conclusion: The plot of seasonal decomposition shows a clear pattern of annual seasonality while the trend-cycle plot shows a strong positive trend of growth in sales over the 5 year period.
This infact confirms the conclusion made under graph number 1.

# Part D: Compute and plot the seasonally adjusted data.


```{r}

# Seasonal adjustment of multiplicative decomposition
p_adj <- p/m$seasonal
autoplot(p_adj)

```

Conclusion: See output above.

# Part E: Change one observation to be an outlier (e.g., add 500 to one observation), and recompute the seasonally adjusted data. What is the effect of the outlier?


```{r}

p[15] <- p[15] + 500
m_1 <- decompose(p, type = "multiplicative")
p_adj_1 <- p/m_1$seasonal
autoplot(p_adj_1)
autoplot(m_1$trend)
autoplot(m_1$random)

```

Conclusion: 
The addition of an outlier to the data impacts decomposition because the classical decomposition methods (multiplicative in this case) are unable to capture these seasonal changes over time. 
The trend-cycle also tends to oversmooth this rapid rise in the data.
This is an inherent weakness of this method.


# Part F: Does it make any difference if the outlier is near the end rather than in the middle of the time series?.

```{r}
p[10] <- p[10] + 500
p[50] <- p[50] + 500
m_2 <- decompose(p, type = "multiplicative")
p_adj_2 <- p/m_2$seasonal
autoplot(p_adj_2)
autoplot(m_2$trend)
autoplot(m_2$random)

```

Conclusion:
The position of the outlier does not seem to make any difference in the decomposition since the classical methods are unable to capture these seasonal changes over time.
Its also evident that the trend-cycle tend to over-smooth the two rapid rises in the data as a result of the two introduced outliers.

